Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-17 doc/invt.tex
small changes

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  E(T)
  =
  E(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot y}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where $\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
%  \label{eq:Gamma_from_G}
%\end{equation}

@@

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimal at
%
\begin{equation}
  t_0 = \frac{ \Gamma } { E(0) }
  ,
  \label{eq:t0_singlebin}
\end{equation}
%
and the minimum is
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the average histogram fluctuation\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:hfluc}).

@@

We will first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s\footnote{This
inequality follows from the nonnegativity of
the quadratic polynomial,
$$
\int_0^T
  dt \sum_{k = 1}^{n-1} \Gamma_k \,
    \left( {\dot y}_k \, \chi - c_k \right)^2
  \equiv
  A_2 \, \chi^2 + A_1 \, \chi + A_0
  ,
$$
which must carry a non-positive discriminant:
$A_1^2 - 4 \, A_0 \, A_2 \le 0$.},
%
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, {\dot y}_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k \, {\dot y}_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 1}^{n-1} \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
Note that the last expression %of Eq. \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err(T)$
given in Eq. \eqref{eq:error_asym}.
%
The equality is achieved
if $\dot y_k\left( q(t) \right) = c_k$
for all $k > 0$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
\note{Derivation.
  Integrating $\dot y_k = c_k$ yields
  \begin{equation}
    y_k(t) = c_k \left(t + t_{k0} \right).
  \tag{UK}
  \label{neq:ui_solution}
  \end{equation}
  Given that $y_k(T) = 1$ and $y_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}
%
Such a solution is possible only if
all nontrivial eigenvalues are equal,
%
%\begin{equation}
  $\lambda_1 = \dots = \lambda_{n-1}$.
%  \notag
%\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$y_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k > 0$.
%
The asymptotic error is then
%
\begin{align}
  \Err_A(T)
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 1 }^{n-1}
    \Gamma_k
  =
  \frac{ \Gamma \, T     }
       { (T + t_0)^2 }
  ,
\label{eq:error_asym_singlebin}
\end{align}
%
where we have used
%
\begin{align}
  \sum_{k=1}^{n-1} \Gamma_k
  &=
  \sum_{t=-\infty}^\infty \sum_{k=1}^{n-1} \langle {\tilde v}_k(t) \, {\tilde v}_k(0) \rangle
  \notag
  \\
  &=
  \sum_{t=-\infty}^\infty \sum_{i=1}^n \rho_i \langle v_{*i}(t) \, v_{*i}(0) \rangle
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  .
  \notag
  %\label{eq:Gamma_G_sum}
\end{align}
%
%
If we further assume $\lambda_1 = \cdots = \lambda_{n-1} = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
supplied with a perfect filter
for the noise modes,
and zero eigenvalues are assigned to the noise modes.
%
For example,
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k < K$,
%
we can construct an updating scheme with
$$
\lambda_0 = \cdots = \lambda_{K-1} = 1,
\mathrm{\; and \;}
\lambda_K = \cdots = \lambda_{n-1} = 0.
$$
%
Then, the updating matrix is, according to
Eq. \eqref{eq:w_from_phi},
%
$$w_{ij} = n \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj},$$
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:xt2_eql}.
%
From Eq. \eqref{eq:xt2_eql}, we have
$$E(0) = \frac{1}{2} a_0 \sum_{k=1}^{n-1} \Gamma_k
= \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
%From Eqs. \eqref{eq:error_tot},
%\eqref{eq:error_res},
%\eqref{eq:tk0},
%and
%\eqref{eq:error_asym_sinc},
%we get
The final error is given by
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 1 }^{ K - 1 }
    \Gamma_k
  ,
\label{eq:error_sinc}
\end{equation}
%
as a generalization of Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme.

@@

%We can use Eq. \eqref{eq:avx_recur} to incrementally
%improve the estimated
%bias potential given by Eq. \eqref{eq:vcorr_equil}.
%
Assuming that we can linearize the histogram correction
in Eq. \eqref{eq:vcorr_equil} as
$\ln x \approx x - 1$ % (for $x \approx 1$),
and apply it to the instantaneous histogram, $h_i(t)$,
then the independent estimate of bias potential in each step
is given by
%
\begin{equation}
  \hat u_i(t) \approx u_i(t) + \frac{ h_i(t) } { \rho_i } - 1
  = u_i(t) + f_i(t) - 1
  .
  \label{eq:uhatt}
\end{equation}
%
Below we shall use capital letter to denote
time average: for variable $x$, we define
\begin{equation}
  X(t) \equiv \frac{1}{t} \sum_{\tau=1}^t x(\tau)
  .
  \label{eq:timeav}
\end{equation}
%
Then after time averaging, Eq. \eqref{eq:uhatt} becomes
\begin{align}
  \hat U_i(t)
  &\approx
  U_i (t)
  + F_i(t) - 1
  \approx
  U_i(t)
  + \ln \frac{ H_i(t) } { \rho_i }
  .
  \label{eq:uhatav}
\end{align}

@@

Here we wish to show that for an FDS simulation
with a nearly-perfect bias potential
(e.g. a long WL simulation under the single-bin updating scheme
with the inverse-time schedule),
the histogram fluctuation
is roughly equal to the final error.
%
With a sufficiently accurate and stable bias potential,
the fluctuation of
the instantaneous histogram is dominated by the random part,
%
\begin{equation}
  f_i(t) \approx \zeta_i(t)
  ,
  \label{eq:xi_zeta}
\end{equation}
and the same holds for the Fourier transform,
$\tilde f_k(t) \approx \tilde \zeta_k(t).$
%
If we define the time average,
%
\begin{align*}
\tilde F_k(T) \equiv \frac 1 T \sum_{t = 1}^T \tilde f_k(t)
=\mathcal F\left[ \frac{ H_i(T) } { \rho_i } \right]_k
,
\end{align*}
%
for the histogram fluctuation of mode $k$,
then the total fluctuation is
%
\begin{align}
G(T)
&\equiv
\sum_{k=1}^{n-1} \tilde F_k^2(T)
=
\sum_{i=1}^{n} \rho_i \, F_{*i}^2(T)
  \notag
  \\
  &=
  \sum_{i=1}^n \rho_i
  \left(
  \frac{H_i(T)}{\rho_i}
  -
  1
  \right)^2
.
  \label{eq:hfluc_def}
\end{align}
From Eq. \eqref{eq:xi_zeta}, we have
%
\begin{align*}
\bigl\langle \tilde F_k^2(T) \bigr\rangle
&=
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
\\
&\approx
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde \zeta_k(t) \, \tilde \zeta_k(0)
\bigr\rangle
=
\frac{ \Gamma_k } { T }
.
\end{align*}
%
The total histogram fluctuation is
\begin{equation}
\bigl\langle G(T) \bigr\rangle
\equiv
\sum_{k=1}^{n-1}
  \bigl\langle
    \tilde F_k^2(T)
  \bigr\rangle
=
\frac{ \Gamma } { T }.
  \label{eq:hfluc}
\end{equation}
%
Comparing this to Eq. \eqref{eq:Emin_singlebin},
we may rephrase the optimality of the inverse-time schedule
as the inequality,
\begin{equation}
  E(T)
  \ge
  \bigl\langle G(T+t_0) \bigr\rangle
  .
  \notag
  %\label{eq:fluc}
\end{equation}
%and the histogram fluctuation
%serves as a good estimate of the final error.
%

@@

Alternatively,
if the propagation along the collective-variable is Markovian,
then in the long time limit, the evolution can be approximately
described by a transition matrix, $\mathbf A$, with
\begin{align*}
\langle h_i(t) \rangle = \rho_i,
\quad
\langle h_i(t) \, h_j(0) \rangle = \bigl(\mathbf A^t\bigr)_{ij} \, \rho_j
\quad \mbox{(for $t \ge 0$)}
.
\end{align*}
Further if the transition matrix satisfies detailed balance
with a set of eigenvectors, $\psi_{li}$,
that are orthonormal with respect to $\pmb\rho$
[cf. Eqs. \eqref{eq:eig_orthonormal_cols} and
\eqref{eq:eig_orthonormal_rows}]
then we have the decomposition
$A_{ij} = \sum_{l=0}^{n-1} \Lambda_l \, \psi_{li} \, \psi_{lj} / \rho_j$,
with $\Lambda_l$ being the eigenvalue of mode $l$.
%
The autocorrelation function can then be computed as
%
\begin{align*}
\left\langle
  \tilde f_k(t) \, \tilde f_k(0)
\right\rangle
=
\sum_{l = 0}^{n-1} \Lambda_l^t \,
\left(
  \sum_{i=1}^n \frac{ \phi_{ki} \, \psi_{li} }{ \rho_i }
\right)^2
.
\end{align*}
%
In particular, if the eigenvectors, $\pmb \phi_k$
and $\pmb \psi_k$, coincide,
we would have
$\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
=
\Lambda_k^t,$
and
\begin{equation}
  \Gamma_k = 1 + 2 \, \sum_{t=1}^\infty \left\langle
    \tilde f_k(t) \, \tilde f_k(0)
  \right\rangle = \frac{1+\Lambda_k}{1-\Lambda_k}
  .
  \label{eq:Gamma_est}
\end{equation}

The transition matrix method usually only for
simple sampling processes without major barrier
along the collective variable, $z$.
%
We may nonetheless estimate the magnitude of $\Gamma_k$
in a few special cases.
%
If sampling is perfect, $\Lambda_k = \delta_{k, 0}$,
then $\Gamma_k = 1$ for $k > 0$.
%
For a local sampling process,
if we borrow \eqref{eq:lambda_Gaussian}
for $\Lambda_k$,
with $\sigma_z$ interpreted as the average move size along $z$,
then
\begin{align}
\Gamma_k = \coth\left[
  \left(
    \frac{ \pi \, k \, \sigma_z } { g \, \Delta z}
  \right)^2
  \right].
\label{eq:Gamma_local}
\end{align}
For small $k$,
we have $\Gamma_k \propto \Delta z^2/(\sigma_z \, k)^2$,\cite{bussi2006}
which means that a doubled sampling range, $\Delta z$,
would make the histogram flattening
at least four times as difficult,
i.e. FDS methods are most effective
for short ranges\cite{wang2001, *wang2001pre}.
%
Using finer bins would, however, not change this value.
%
For a modestly large $k$,
$\Gamma_k \approx 1$.
and by Eq. \eqref{eq:xt2_eql},
we would expect a roughly flat tail
for the single-bin updating scheme
as $\langle \tilde v_k^2 \rangle \propto \Gamma_k$,
cf. Fig. \ref{fig:lj_xerr}.

@@

For $n \gg 1$,
we define an equivalent angular variable, $\varphi$,
over $[0, 2 \, \pi/g]$
with bin size
$\delta \varphi = 2 \, \pi/(g \, n)$.
The variable is related to the bin index, $l$, as
$\varphi = l \, \delta \varphi$,
and
$\mu_l = \mu(\varphi) \, \delta \varphi$.
%
By assuming the largest bin cutoff, $b \approx g \, n/2$,
we can approximate Eq. \eqref{eq:wband_eigenvalue}
as an integral:
%
\begin{equation}
  \lambda_k
  =
  2 \int_0^\pi
    \mu(\varphi) \, \cos (k \, \varphi) \, d\varphi,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
  1 = 2 \int_0^\pi \mu(\varphi) \, d\varphi.
\label{eq:mu_int}
\end{equation}
\note{
\begin{align*}
  \lambda_k
  &\approx
  \int_0^b \mu_l \, \cos(k \, \varphi) \, dl
  \\
  &=
  \int_0^b \mu(\varphi) \, \cos(k \, \varphi) \, \delta \varphi \, dl
  \\
  &=
  \int_0^{b \, \delta \varphi} \mu(\varphi) \, \cos(k \, \varphi) \, d\varphi
\end{align*}
The upper limit,
$b \, \delta \varphi \approx (g\,n/2) \, [2\,\pi/(g\,n)] = \pi$.
}

>>>>

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  \Err(T)
  =
  \Err(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot y}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where $\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
%  \label{eq:Gamma_from_G}
%\end{equation}

@@

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$\Err(T) = [ \Err(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimal at
%
\begin{equation}
  t_0 = \frac{ \Gamma } { \Err(0) }
  ,
  \notag
  %\label{eq:t0_singlebin}
\end{equation}
%
and the minimum is
%
\begin{equation}
  \Err(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the average histogram fluctuation\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:hfluc}).

@@

We will first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s\footnote{This
inequality follows from the nonnegativity of
the quadratic polynomial,
$$
\int_0^T
  dt \sum_{k = 1}^{n-1} \Gamma_k \,
    \left( {\dot y}_k \, \chi - c_k \right)^2
  \equiv
  A_2 \, \chi^2 + A_1 \, \chi + A_0
  ,
$$
which must carry a non-positive discriminant:
$A_1^2 - 4 \, A_0 \, A_2 \le 0$.},
%
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, {\dot y}_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k \, {\dot y}_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 1}^{n-1} \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\notag
%\label{eq:CSineq}
\end{align}
%
Note that the last expression %of Eq. \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err(T)$
given in Eq. \eqref{eq:error_asym}.
%
The equality is achieved
if $\dot y_k\left( q(t) \right) = c_k$
for all $k > 0$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
\note{Derivation.
  Integrating $\dot y_k = c_k$ yields
  \begin{equation}
    y_k(t) = c_k \left(t + t_{k0} \right).
  \tag{UK}
  \label{neq:ui_solution}
  \end{equation}
  Given that $y_k(T) = 1$ and $y_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}
%
Such a solution is possible only if
all nontrivial eigenvalues are equal,
%
%\begin{equation}
  $\lambda_1 = \dots = \lambda_{n-1}$.
%  \notag
%\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$y_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k > 0$.
%
The asymptotic error is then
%
\begin{align}
  \Err_A(T)
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 1 }^{n-1}
    \Gamma_k
  =
  \frac{ \Gamma \, T     }
       { (T + t_0)^2 }
  ,
\label{eq:error_asym_singlebin}
\end{align}
%
where we have used
%
\begin{align}
  \sum_{k=1}^{n-1} \Gamma_k
  &=
  \sum_{t=-\infty}^\infty \sum_{k=1}^{n-1} \langle {\tilde v}_k(t) \, {\tilde v}_k(0) \rangle
  \notag
  \\
  &=
  \sum_{t=-\infty}^\infty \sum_{i=1}^n \rho_i \langle v_{*i}(t) \, v_{*i}(0) \rangle
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  .
  \notag
  %\label{eq:Gamma_G_sum}
\end{align}
%
%
If we further assume $\lambda_1 = \cdots = \lambda_{n-1} = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
A bandpass updating scheme is essentially
the single-bin updating scheme
supplied with a perfect filter
for the noise modes,
and zero eigenvalues are assigned to the noise modes.
%
For example,
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k < K$,
%
we can construct an updating scheme with
$$
\lambda_0 = \cdots = \lambda_{K-1} = 1,
\mathrm{\; and \;}
\lambda_K = \cdots = \lambda_{n-1} = 0.
$$
%
Then, the updating matrix is, according to
Eq. \eqref{eq:w_from_phi},
%
$$w_{ij} = n \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj},$$
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers $t_0 = \Gamma/\Err(0)$ %Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:xt2_eql}.
%
From Eq. \eqref{eq:xt2_eql}, we have
$$\Err(0) = \frac{1}{2} a_0 \sum_{k=1}^{n-1} \Gamma_k
= \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
%From Eqs. \eqref{eq:error_tot},
%\eqref{eq:error_res},
%\eqref{eq:tk0},
%and
%\eqref{eq:error_asym_sinc},
%we get
The final error is given by
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 1 }^{ K - 1 }
    \Gamma_k
  ,
  \notag
  %\label{eq:error_sinc}
\end{equation}
%
as a generalization of Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme.

@@

%We can use Eq. \eqref{eq:avx_recur} to incrementally
%improve the estimated
%bias potential given by Eq. \eqref{eq:vcorr_equil}.
%
Assuming that we can linearize the histogram correction
in Eq. \eqref{eq:vcorr_equil} as
$\ln x \approx x - 1$ % (for $x \approx 1$),
and apply it to the instantaneous histogram, $h_i(t)$,
then the independent estimate of bias potential in each step
is given by
%
\begin{equation}
  \hat u_i(t) \approx u_i(t) + \frac{ h_i(t) } { \rho_i } - 1
  = u_i(t) + f_i(t) - 1
  .
  \label{eq:uhatt}
\end{equation}
%
Below we shall use capital letter to denote
time average: for variable $x$, we define
\begin{equation}
  X(t) \equiv \frac{1}{t} \sum_{\tau=1}^t x(\tau)
  .
  \notag
  %\label{eq:timeav}
\end{equation}
%
Then after time averaging, Eq. \eqref{eq:uhatt} becomes
\begin{align}
  \hat U_i(t)
  &\approx
  U_i (t)
  + F_i(t) - 1
  \approx
  U_i(t)
  + \ln \frac{ H_i(t) } { \rho_i }
  .
  \label{eq:uhatav}
\end{align}

@@

Here we wish to show that for an FDS simulation
with a nearly-perfect bias potential
(e.g. a long WL simulation under the single-bin updating scheme
with the inverse-time schedule),
the histogram fluctuation
is roughly equal to the final error.
%
With a sufficiently accurate and stable bias potential,
the fluctuation of
the instantaneous histogram is dominated by the random part,
%
\begin{equation}
  f_i(t) \approx \zeta_i(t)
  ,
  \label{eq:xi_zeta}
\end{equation}
and the same holds for the Fourier transform,
$\tilde f_k(t) \approx \tilde \zeta_k(t).$
%
If we define the time average,
%
\begin{align*}
\tilde F_k(T) \equiv \frac 1 T \sum_{t = 1}^T \tilde f_k(t)
=\mathcal F\left[ \frac{ H_i(T) } { \rho_i } \right]_k
,
\end{align*}
%
for the histogram fluctuation of mode $k$,
then the total fluctuation is
%
\begin{align}
G(T)
&\equiv
\sum_{k=1}^{n-1} \tilde F_k^2(T)
=
\sum_{i=1}^{n} \rho_i \, F_{*i}^2(T)
  \notag
  \\
  &=
  \sum_{i=1}^n \rho_i
  \left(
  \frac{H_i(T)}{\rho_i}
  -
  1
  \right)^2
.
  \notag
  %\label{eq:hfluc_def}
\end{align}
From Eq. \eqref{eq:xi_zeta}, we have
%
\begin{align*}
\bigl\langle \tilde F_k^2(T) \bigr\rangle
&=
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
\\
&\approx
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde \zeta_k(t) \, \tilde \zeta_k(0)
\bigr\rangle
=
\frac{ \Gamma_k } { T }
.
\end{align*}
%
The total histogram fluctuation is
\begin{equation}
\bigl\langle G(T) \bigr\rangle
\equiv
\sum_{k=1}^{n-1}
  \bigl\langle
    \tilde F_k^2(T)
  \bigr\rangle
=
\frac{ \Gamma } { T }
  .
  \notag
  %\label{eq:hfluc}
\end{equation}
%
Comparing this to Eq. \eqref{eq:Emin_singlebin},
we may rephrase the optimality of the inverse-time schedule
as the inequality,
\begin{equation}
  \Err(T)
  \ge
  \bigl\langle G(T+t_0) \bigr\rangle
  .
  \notag
  %\label{eq:fluc}
\end{equation}
%and the histogram fluctuation
%serves as a good estimate of the final error.
%

@@

Alternatively,
if the propagation along the collective-variable is Markovian,
then in the long time limit, the evolution can be approximately
described by a transition matrix, $\mathbf A$, with
\begin{align*}
\langle h_i(t) \rangle = \rho_i,
\quad
\langle h_i(t) \, h_j(0) \rangle = \bigl(\mathbf A^t\bigr)_{ij} \, \rho_j
\quad \mbox{(for $t \ge 0$)}
.
\end{align*}
Further if the transition matrix satisfies detailed balance
with a set of eigenvectors, $\psi_{li}$,
that are orthonormal with respect to $\pmb\rho$
[cf. Eqs. \eqref{eq:eig_orthonormal_cols} and
\eqref{eq:eig_orthonormal_rows}]
then we have the decomposition
$A_{ij} = \sum_{l=0}^{n-1} \Lambda_l \, \psi_{li} \, \psi_{lj} / \rho_j$,
with $\Lambda_l$ being the eigenvalue of mode $l$.
%
The autocorrelation function can then be computed as
%
\begin{align*}
\left\langle
  \tilde f_k(t) \, \tilde f_k(0)
\right\rangle
=
\sum_{l = 0}^{n-1} \Lambda_l^t \,
\left(
  \sum_{i=1}^n \frac{ \phi_{ki} \, \psi_{li} }{ \rho_i }
\right)^2
.
\end{align*}
%
In particular, if the eigenvectors, $\pmb \phi_k$
and $\pmb \psi_k$, coincide,
we would have
$\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
=
\Lambda_k^t,$
and
\begin{equation}
  \Gamma_k = 1 + 2 \, \sum_{t=1}^\infty \left\langle
    \tilde f_k(t) \, \tilde f_k(0)
  \right\rangle = \frac{1+\Lambda_k}{1-\Lambda_k}
  .
  \notag
  %\label{eq:Gamma_est}
\end{equation}

The transition matrix method usually works only for
simple sampling processes without major barrier
along the collective variable, $z$.
%
We may nonetheless estimate the magnitude of $\Gamma_k$
in a few special cases.
%
If sampling is perfect, $\Lambda_k = \delta_{k, 0}$,
then $\Gamma_k = 1$ for $k > 0$.
%
For a local sampling process,
if we borrow \eqref{eq:lambda_Gaussian}
for $\Lambda_k$,
with $\sigma_z$ interpreted as the average move size along $z$,
then
\begin{align}
\Gamma_k = \coth\left[
  \left(
    \frac{ \pi \, k \, \sigma_z } { g \, \Delta z}
  \right)^2
  \right].
\label{eq:Gamma_local}
\end{align}
For small $k$,
we have $\Gamma_k \propto \Delta z^2/(\sigma_z \, k)^2$,\cite{bussi2006}
which means that a doubled sampling range, $\Delta z$,
would make the histogram flattening
at least four times as difficult,
i.e. FDS methods are most effective
for short ranges\cite{wang2001, *wang2001pre}.
%
Using finer bins would, however, not change this value.
%
For a modestly large $k$,
$\Gamma_k \approx 1$.
and by Eq. \eqref{eq:xt2_eql},
we would expect a roughly flat tail
for the single-bin updating scheme
as $\langle \tilde v_k^2 \rangle \propto \Gamma_k$,
cf. Fig. \ref{fig:lj_xerr}.

@@

For $n \gg 1$,
we define an equivalent angular variable, $\varphi$,
over $[0, 2 \, \pi/g]$
with bin size
$\delta \varphi = 2 \, \pi/(g \, n)$.
The variable is related to the bin index, $l$, as
$\varphi = l \, \delta \varphi$,
and
$\mu_l = \mu(\varphi) \, \delta \varphi$.
%
By assuming the largest bin cutoff, $b \approx g \, n/2$,
we can approximate Eq. \eqref{eq:wband_eigenvalue}
as an integral:
%
\begin{equation}
  \lambda_k
  =
  2 \int_0^\pi
    \mu(\varphi) \, \cos (k \, \varphi) \, d\varphi,
  \notag
  %\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
  1 = 2 \int_0^\pi \mu(\varphi) \, d\varphi.
\label{eq:mu_int}
\end{equation}
\note{
\begin{align*}
  \lambda_k
  &\approx
  \int_0^b \mu_l \, \cos(k \, \varphi) \, dl
  \\
  &=
  \int_0^b \mu(\varphi) \, \cos(k \, \varphi) \, \delta \varphi \, dl
  \\
  &=
  \int_0^{b \, \delta \varphi} \mu(\varphi) \, \cos(k \, \varphi) \, d\varphi
\end{align*}
The upper limit,
$b \, \delta \varphi \approx (g\,n/2) \, [2\,\pi/(g\,n)] = \pi$.
}
